Covariate analysis of MSSM reprocessed counts (with CQN normalisation and covariates adjusted)

### 2017-08-11 code/Rmd/MSSM_REPROCESSED.Rmd
Added ApoE differential expression to MSSM processing markdown

METADATA = METADATA %>%
  tidyr::unite(BrainRegion.Diagnosis, BrainRegion, Diagnosis, sep = '.')

@@

METADATA = METADATA %>%
  dplyr::filter(RIN >= 5)
```

@@

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering}
primaryVariable <- c("CDR", "BrainRegion.Diagnosis")
FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrainRegion.Diagnosis", "CDR")
ContCovariates <- c("RIN", "RIN2", "AOD","PMI",  "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")

@@

Distribution of samples are: `r kable(tmp)`
### Winsorise counts (Gene outliers in samples)
```{r winsorise.data}
LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)

@@

  x[x < (mn-3*std.dev)] = x[ind.low]
  x[x > (mn+3*std.dev)] = x[ind.high]
  return(x)
}) %>% t
CQN.GENE_EXPRESSION$E = LOG.CPM

PROCESSED_COUNTS$filteredExprMatrix$counts = NEW.COUNTS
```

@@

  # Estimate voom weights for dispersion control
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)

  # Calculate correlation values of random effects
  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
                                             block = COVARIATES$individualIdentifier)

  # Re-estimate voom weights
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                              design=DM1$design, plot=F, 
                              block = COVARIATES$individualIdentifier, 
                              correlation = correlation$cor)

@@

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
  print(loopCount)
}
RESIDUAL.GENE_EXPRESSION
modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

@@

@@

@@

@@

Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
```{r diff.exp2, fig.height=15, fig.width=13}
## Modeling Diagnosis, msex and age_death conjointly
COVARIATES$BrainRegion.Diagnosis.SEX = paste(COVARIATES$BrainRegion.Diagnosis, 
                                             factor(COVARIATES$SEX, labels = c('F' = 'FEMALE', 'M' = 'MALE')), 
                                             sep = '.') %>% factor

@@

Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
```{r diff.exp3, fig.height=15, fig.width=13}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[2], setdiff(postAdjustCovars, 'AOD')), drop = F], Intercept = F)
ind = grep('Diagnosis', colnames(DESIGN$design))
DESIGN$design[,ind] = DESIGN$design[,ind] * COVARIATES[rownames(DESIGN$design),'AOD']
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
colnames(DESIGN$design)[ind] = gsub('BrainRegion.Diagnosis', '', paste0(colnames(DESIGN$design)[ind], '.AOD'))

@@

# Estimate voom weights for dispersion control
VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design=cbind(MODEL1, SURR.VAR), plot=F)

@@

# Re-estimate voom weights
VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design = cbind(MODEL1, SURR.VAR), plot=F,
                        block = COVARIATES$individualIdentifier, 
                        correlation = correlation$cor)

@@

# Estimate voom weights for dispersion control
VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design=DESIGN.NET, plot=F)

@@

# Re-estimate voom weights
VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design = DESIGN.NET, plot=F,
                        block = COVARIATES$individualIdentifier, 
                        correlation = correlation$cor)

@@

# Function to change entity ACL (also creates new acls if it doesn't exist)
synChangeEntityACL <- function(id, new.racl){
  tryCatch({ 
    # If old acl exist update
    old.acl = synGetEntityACL(id)
    old.acl@resourceAccess = new.racl
    synUpdateEntityACL(old.acl)
  }, error = function(e){ 
    # If not create a new acl
    new.acl = AccessControlList(id = id, resourceAccess = new.racl)
    synCreateEntityACL(new.acl)
  })
}

@@

# Store logCPM
CQN.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

@@

# Delete folder acl
synDeleteEntityACL(CODE@properties$id)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

>>>>

METADATA = METADATA %>%
  dplyr::mutate(BrainRegion1 = BrainRegion) %>%
  tidyr::unite(BrainRegion.Diagnosis, BrainRegion, Diagnosis, sep = '.')

@@

METADATA = METADATA %>%
  dplyr::filter(RIN >= 5)

METADATA$ApoE[METADATA$Apo1 != 4 & METADATA$Apo2 != 4] = 0
METADATA$ApoE[METADATA$Apo1 == 4 | METADATA$Apo2 == 4] = 1
METADATA$ApoE[METADATA$Apo1 == 4 & METADATA$Apo2 == 4] = 2
METADATA$BrainRegion.ApoE = paste(METADATA$BrainRegion1, METADATA$ApoE, sep = '.') %>%
  factor()
```

@@

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering}
primaryVariable <- c("CDR", "BrainRegion.Diagnosis", "BrainRegion.ApoE")
FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrainRegion.Diagnosis", "CDR", "BrainRegion.ApoE")
ContCovariates <- c("RIN", "RIN2", "AOD","PMI",  "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")

@@

Distribution of samples are: `r kable(tmp)`

### Winsorise counts (Gene outliers in samples)
```{r winsorise.data}
LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)

@@

  x[x < (mn-3*std.dev)] = x[ind.low]
  x[x > (mn+3*std.dev)] = x[ind.high]
  return(x)
}) %>% t
CQN.GENE_EXPRESSION$E = LOG.CPM
NEW.COUNTS = (2^LOG.CPM) * t(replicate(dim(LOG.CPM)[1], LIB.SIZE))/1e6
```

@@

  # Estimate voom weights for dispersion control
  VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, design=DM1$design, plot=F)

  # # Calculate correlation values of random effects
  correlation = parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION,
                                             block = COVARIATES$individualIdentifier)
  # correlation = list(); correlation$cor = 0.4

  # Re-estimate voom weights
  VOOM.GENE_EXPRESSION = voom(NEW.COUNTS, 
                              design=DM1$design, plot=F, 
                              block = COVARIATES$individualIdentifier, 
                              correlation = correlation$cor)

@@

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
  # print(loopCount)
}
modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

@@

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)

@@

# Re-estimate voom weights
VOOM.WEIGHTS = voom(NEW.COUNTS,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$individualIdentifier, 
                    correlation = correlation$cor)

@@

### Differential expression analysis (with BrainRegion.ApoE as primary variable)
Differential expression is performed on the primary variable by controlling for covariates identified above

Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp2, fig.height=15, fig.width=13}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[3], postAdjustCovars), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)

# Re-estimate voom weights
VOOM.WEIGHTS = voom(NEW.COUNTS,
                    design=DESIGN$design, plot=F,
                    block = COVARIATES$individualIdentifier, 
                    correlation = correlation$cor)

# Fit contrast
cntr = c()
for (region in c('FP', 'STG', 'PHG', 'IFG')){
  tmp = combn(paste0('BrainRegion.ApoE', region, '.', c('0', '1', '2')),2)
  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
}

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.ApoE','',Comparison),
                Study = 'MSSM',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('BrainRegion1','against'), sep = '\\.') %>%
  dplyr::mutate(Comparison = paste(reference, against, sep = '-')) %>%
  dplyr::select(-BrainRegion1) %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

all.diff.exp = c(all.diff.exp, list(ApoE = DE))
all.fit = c(all.fit, list(ApoE = FIT))
```

@@

Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
```{r diff.exp3, fig.height=15, fig.width=13}
## Modeling Diagnosis, msex and age_death conjointly
COVARIATES$BrainRegion.Diagnosis.SEX = paste(COVARIATES$BrainRegion.Diagnosis, 
                                             factor(COVARIATES$SEX, labels = c('F' = 'FEMALE', 'M' = 'MALE')), 
                                             sep = '.') %>% factor

@@

Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
```{r diff.exp4, fig.height=15, fig.width=13}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[2], setdiff(postAdjustCovars, 'AOD')), drop = F], Intercept = F)
ind = grep('Diagnosis', colnames(DESIGN$design))
DESIGN$design[,ind] = DESIGN$design[,ind] * COVARIATES[rownames(DESIGN$design),'AOD']
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]
colnames(DESIGN$design)[ind] = gsub('BrainRegion.Diagnosis', '', paste0(colnames(DESIGN$design)[ind], '.AOD'))

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(NEW.COUNTS, design=DESIGN$design, plot=F)

@@

# Estimate voom weights for dispersion control
VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
                        design=cbind(MODEL1, SURR.VAR), plot=F)

@@

# Re-estimate voom weights
VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
                        design = cbind(MODEL1, SURR.VAR), plot=F,
                        block = COVARIATES$individualIdentifier, 
                        correlation = correlation$cor)

@@

# Estimate voom weights for dispersion control
VOOM.NET.WEIGHTS = voom(NEW.COUNTS, 
                        design=DESIGN.NET, plot=F)

@@

# Re-estimate voom weights
VOOM.NET.WEIGHTS = voom(NEW.COUNTS, 
                        design = DESIGN.NET, plot=F,
                        block = COVARIATES$individualIdentifier, 
                        correlation = correlation$cor)

@@

@@

# Store filtered counts
NEW.COUNTS %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_eCounts.tsv', sep = '\t', row.names=F, quote=F)
ECOUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_eCounts.tsv', name = 'Counts (estimated)', parentId = CODE$properties$id)
annotations(ECOUNT_OBJ) = annotations(CODE)
annotations(ECOUNT_OBJ)$dataSubType = 'estimatedCounts'
ECOUNT_OBJ = synStore(ECOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store logCPM
CQN.GENE_EXPRESSION$y %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'MSSM_FP_STG_PHG_IFG_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('MSSM_FP_STG_PHG_IFG_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

@@

# Delete folder acl
synDeleteEntityACL(CODE@properties$id)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (estimated)                         |  `r ECOUNT_OBJ$properties$id`    |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
